---
title: "cytoeffect Workflow"
author: "Christof Seiler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cytoeffect Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 5
)
```

In this vignette, we show the `cytoeffect` workflow for both the full Bayesian hierarchical model and the parametric bootstrap with the simplified model.

## Generate Data

Simulate dataset.

```{r}
library("tibble")
library("magrittr")
library("stringr")
library("MASS")
library("Matrix")
library("dplyr")

# parameters
n_markers = 5            # number of markers
n_true = 1               # number of differential markers
n_samples = 8            # donors per group
paired = TRUE            # paired experimental design
n_cells = 100            # per sample
pi = 0.05                # zero inflation probability
beta_treatment = log(20) # log of mean in treatment
beta_control = log(5)    # log of mean in control
rho_b = 0.5              # cell level correlation
rho_u = 0.5              # donor level correlation
seed = 0xdada            # set random seed

# simulation parameters
set.seed(seed)

if (paired) {
  n_donors = n_samples / 2
} else {
  n_donors = n_samples
}

# patient information
donor = rep(1:n_donors, each = n_cells)
if (paired) {
  donor %<>% rep(2)
  condition = c(rep("treatment", length(donor) / 2),
                rep("control", length(donor) / 2))
} else {
  condition = c(rep("treatment", length(donor) / 2),
                rep("control", length(donor) / 2))
}
df = tibble(donor, condition)

# generate protein counts
protein_names = paste0("m", str_pad(1:n_markers, width = 2, pad = "0"))
rcorr = function(rho) {
  corr = rho ^ toeplitz(0:(n_markers - 1))
  # random positive and negative correlations
  lower_tri = lower.tri(corr)
  upper_tri = upper.tri(corr)
  signs = sample(c(1, -1), size = sum(lower_tri), replace = TRUE)
  corr_sign = diag(1, n_markers)
  corr_sign[lower_tri] = signs * corr[lower_tri]
  corr_sign[upper_tri] = t(corr_sign)[upper_tri]
  # project to nearest positive definite matrix
  as.matrix(nearPD(corr_sign, corr = TRUE)$mat)
}
Sigma_b = rcorr(rho_b) # cell level variability
Sigma_u = rcorr(rho_u) # donor level variability
b = mvrnorm(n = nrow(df), mu = rep(0, n_markers), Sigma_b)
u = mvrnorm(n = n_donors, mu = rep(0, n_markers), Sigma_u)
u = u[donor,]
beta = matrix(beta_control, nrow = nrow(b), ncol = n_markers)
# collider confounding
beta[, 1:n_true] = ifelse(condition == "treatment", beta_treatment, beta_control)
log_lambda = beta + b + u
lambda = exp(log_lambda)
y = rpois(length(lambda), lambda)
dim(y) = dim(lambda)
I = rbinom(n = length(y), size = 1, prob = pi)
I = matrix(I, ncol = n_markers)
y = y * (1 - I)
colnames(y) = protein_names
df %<>% bind_cols(as_tibble(y))
df$condition %<>% factor(levels = c("control",
                                    "treatment"))
df %<>% mutate(donor = paste0("pid", str_pad(donor, width = 2, pad = "0")))
```

## Full Bayesian Model
 
Sample from posterior distribution using Stan.

```{r}
library("cytoeffect")
fit = poisson_lognormal(df, protein_names, condition = "condition", group = "donor",
                        warmup = 100, iter = 200, r_donor = 2, num_chains = 1)
```

Plot marginal credible intervals.

```{r}
plot(fit, type = "theta")
plot(fit, type = "beta")
plot(fit, type = "sigma")
plot(fit, type = "Cor")
```

Multivariate DiSTATIS plot.

```{r}
plot_distatis(fit)
```

## Parametric Bootstrap with Simplified Model

Fit model using composite maximum likelihood estimatation.

```{r}
fit = poisson_lognormal_mcle(df, protein_names, condition = "condition", group = "donor",
                             ncores = 1)
```

Plot marginal credible intervals.

```{r}
plot(fit, type = "beta")
plot(fit, type = "sigma")
plot(fit, type = "Cor")
```

Multivariate DiSTATIS plot.

```{r}
plot_distatis(fit)
```

Session info.

```{r}
sessionInfo()
```
