---
title: "cytoeffect Workflow"
author: "Christof Seiler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cytoeffect Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 5
)
library("cytoeffect")
```

In this vignette, we show the `cytoeffect` workflow for both the full Bayesian hierarchical model and the parametric bootstrap with the simplified model.

## Generate Data

Simulate dataset.

```{r}
df = simulate_data()
str(df)
df
condition = "condition"
group = "donor"
protein_names = names(df)[3:ncol(df)]
```

## Full Bayesian Model
 
Sample from posterior distribution using Stan.

```{r}
fit = poisson_lognormal(df, 
                        protein_names = protein_names, 
                        condition = condition, 
                        group = group,
                        r_donor = 2, 
                        warmup = 100, iter = 200, 
                        num_chains = 1)
```

Plot marginal credible intervals.

```{r}
plot(fit, type = "theta")
plot(fit, type = "beta")
plot(fit, type = "sigma")
plot(fit, type = "Cor")
```

Multivariate DiSTATIS plot.

```{r}
plot_distatis(fit)
```

## Parametric Bootstrap with Simplified Model

Fit model using composite maximum likelihood estimatation.

```{r}
fit = poisson_lognormal_mcle(df, 
                             protein_names = protein_names, 
                             condition = condition, 
                             group = group,
                             ncores = 1)
```

Plot marginal credible intervals.

```{r}
plot(fit, type = "beta")
plot(fit, type = "sigma")
plot(fit, type = "Cor")
```

Multivariate DiSTATIS plot.

```{r}
plot_distatis(fit)
```
