---
title: "Poisson Log-Normal Model"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: bibliography.bib
params:
  ncells: "Inf"
  seed: "0xdada"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

Reanalysis of mass cytometry data from @aghaeepour2017immune using the Poisson Log-Normal Model.

# Prerequisites

Parse input parameters.

```{r parse_input_parameters}
ncells = floor(as.numeric(params$ncells))
ncells
seed = params$seed
seed
ncores = 8
```

Install packages.

```{r install_packages, warning=FALSE, message=FALSE, eval=FALSE}
pkgs_needed = c("devtools","tidyverse","magrittr","SummarizedExperiment",
                "ggthemes","cowplot","RColorBrewer","broom","hexbin",
                "intergraph","igraph","ggnetwork","ggcorrplot","MASS",
                "parallel","dplyr")
letsinstall = setdiff(pkgs_needed, installed.packages())
if (length(letsinstall) > 0) {
  BiocManager::install(letsinstall)
}
devtools::install_github("ChristofSeiler/cytoeffect")
```

Load packages.

```{r load_packages}
library("cytoeffect")
library("tidyverse")
library("magrittr")
library("SummarizedExperiment")
library("ggthemes")
library("cowplot")
library("RColorBrewer")
library("broom")
library("intergraph")
library("igraph")
library("ggnetwork")
library("ggcorrplot")
library("MASS")
library("parallel")
library("dplyr")
theme_set(theme_few())
```

# Load Data

Load ``SummarizedExperiment`` object from CytoGLMM workflow.

```{r load_summarized_experiment}
load("se_aghaeepour2017immune.Rdata")
exprs = assay(se_aghaeepour2017immune)
sample_info = rowData(se_aghaeepour2017immune)
sample_info_names = names(sample_info)
df_samples = cbind(as.data.frame(exprs), as.data.frame(sample_info))
df_samples %<>% as.tibble
protein_names = colData(se_aghaeepour2017immune) %>% 
  as.data.frame %>% 
  dplyr::filter(type == "function") %>%
  .$protein_name
gate_protein_names = colData(se_aghaeepour2017immune) %>% 
  as.data.frame %>% 
  dplyr::filter(type == "phenotype") %>%
  .$protein_name
```

# Fit Model

Cells count of subsets.

```{r cell_count}
df_samples %>% group_by(term,celltype,donor) %>% tally %>% arrange(n)
df_samples %>% group_by(term,celltype,donor) %>% tally %>% arrange(desc(n))
```

## Subsampling

Subsample to only one cell type and one donor.

```{r one_donor_one_cell_subset}
df_samples_subset = df_samples %>% dplyr::filter(celltype == "NK")
df_samples_subset %<>% dplyr::select(protein_names,gate_protein_names,sample_info_names)
```

Count gating proteins.

```{r count_gating_proteins}
df_samples_subset %<>% 
  mutate_(cell_size = paste(gate_protein_names,collapse = "+"))
ggplot(df_samples_subset, aes(cell_size)) + 
  geom_histogram(bins = 100) + 
  facet_wrap(~term, ncol = 1)
```

Explore possible NK cells heterogeneity as a function of CD16 and CD56 expression levels.

```{r gating_markers}
tfm = function(x) asinh(x/5)
colorscale = scale_fill_gradientn(
  colors = rev(brewer.pal(9, "YlGnBu")), 
  values = c(0, exp(seq(-5, 0, length.out = 100)))
)
gate_nk = c("CD16","CD56")
df_gate_nk = df_samples_subset %>% dplyr::select(donor, file_name, term, gate_nk)
df_gate_nk %<>% mutate_at(gate_nk, tfm)
ggplot(df_gate_nk, aes(x = CD16, y = CD56)) +
  geom_hex(bins = 16) +
  coord_fixed() +
  facet_wrap(~term) + 
  colorscale
ggplot(df_gate_nk, aes(x = CD16, y = CD56)) +
  geom_hex(bins = 16) +
  coord_fixed() +
  facet_wrap(~donor, ncol = 6) + 
  colorscale
ggplot(df_gate_nk %>% 
         gather(protein_name, value, -donor, -file_name, -term), 
       aes(value, fill = term)) + 
  geom_histogram(bins = 30, alpha= 0.5, position = "identity") +
  facet_wrap(~protein_name) +
  scale_color_few()
```

NK subset analysis.

```{r nk_subset_analysis}
df_samples_subset %<>% mutate(CD56_tfm = tfm(CD56))
df_samples_subset %<>% mutate(CD56_tfm = (CD56_tfm-mean(CD56_tfm))/sd(CD56_tfm))
df_samples_subset %<>% mutate(CD16_tfm = tfm(CD16))
df_samples_subset %<>% mutate(CD16_tfm = (CD16_tfm-mean(CD16_tfm))/sd(CD16_tfm))
tb_fit = lapply(protein_names, function(y) {
  fit = glm(paste(y, "~ term + CD56_tfm + CD16_tfm"), data = df_samples_subset)
  tidy(fit) %>% 
    add_column(protein_name = y)
  }) %>% bind_rows()
ggplot(tb_fit, # %>% dplyr::filter(term != "(Intercept)"),
       aes(x = estimate, y = protein_name)) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_point() + 
  geom_errorbarh(aes(xmax = estimate + 1.96*std.error, 
                     xmin = estimate - 1.96*std.error)) +
  facet_wrap(~term)
```

Subsample cells so that to a maximum number of cells per celltype and donor.

```{r subsample_cells}
if(nrow(df_samples_subset) > ncells) {
  set.seed(seed)
  # subsample depending on max cell count
  df_count = df_samples_subset %>% group_by(donor) %>% tally() %>%
    mutate(nnew = ifelse(n > ncells,ncells,n))
  # create table with a data frame in one column
  df_nested = df_samples_subset %>% group_by(donor) %>% nest() %>%
    left_join(df_count,by = "donor")
  # subsample per donor
  df_samples_subset = df_nested %>%
    mutate(samp = map2(data, nnew, sample_n)) %>%
    dplyr::select(donor, samp) %>%
    unnest()
}
```

Subset cells count.

```{r cell_count_subsample}
df_samples_subset %>% group_by(term,celltype,donor) %>% tally %>% 
  arrange(n)
df_samples_subset %>% group_by(term,celltype,donor) %>% tally %>% 
  arrange(desc(n))
```

## HMC Sampling

Sample from posterior distribution (initialization details in paper).

```{r set_reference}
df_samples_subset$term %<>% factor(levels = c("1st trimester",
                                              "3rd trimester"))
```

```{r poisson_sampling_load, echo=FALSE, eval=TRUE}
load(file = paste0("cytoeffect_poisson_ncells",ncells,".Rdata"))
```

```{r poisson_sampling_run, eval=FALSE}
obj = cytoeffect::poisson_lognormal(df_samples_subset, protein_names, 
                                    condition = "term", group = "donor",
                                    iter = 82, warmup = 50,
                                    num_chains = ncores)
save(obj,file = paste0("cytoeffect_poisson_ncells",ncells,".Rdata"))
```

Postprocessing of posterior samples. Traceplot of posterior samples.

```{r post_sampling, fig.wide=TRUE}
pars_str = "beta"
rstan::traceplot(obj$fit_mcmc, inc_warmup = FALSE, pars = pars_str)
```

Some more MCMC diagnostics. According to empirically findings, Rhat > 1.1 is usually indicative of problems in the fit.

```{r mcmc_diagnostics}
pars = c("beta",
         "sigma","sigma_term","sigma_donor",
         "Cor","Cor_term","Cor_donor")
tb = summary(obj$fit_mcmc, 
             pars = pars)$summary %>% 
  as.tibble(rownames = "pars", .before = 1) %>% 
  dplyr::select(pars, n_eff, Rhat)
tb %<>% na.omit() # Stan fills upper triangle with zeros
tb %>% arrange(n_eff)
tb %>% arrange(desc(Rhat))
tb %>% summarize(min = min(n_eff), max = max(n_eff))
tb %>% summarize(min = min(Rhat), max = max(Rhat))
```

# Results

Plot posterior regression coefficients.

```{r plot_beta}
p1 = plot(obj, type = "beta") + 
  ggtitle(expression("Fixed Effects"~beta)) +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(ncol = 1)) + 
  scale_color_few()
p1
plot(obj, type = "beta") + 
  facet_wrap(~condition, scales = "free_x") +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(ncol = 1)) + 
  scale_color_few()
```

Extract expected count difference for pSTAT1.

```{r fixed_effects_pSTAT1}
post_beta = rstan::extract(obj$fit_mcmc, pars = "beta")[[1]]
first_index = which(obj$conditions == "1st trimester")
third_index = which(obj$conditions == "3rd trimester")
pstat1_index = which(obj$protein_names == "pSTAT1")
first_log_count = quantile(post_beta[,pstat1_index,first_index], 
                           probs = c(0.025, 0.5, 0.975))
first_log_count
exp(first_log_count)
third_log_count = quantile(post_beta[,pstat1_index,third_index], 
                           probs = c(0.025, 0.5, 0.975))
third_log_count
exp(third_log_count)
diff_log_count = quantile(
  post_beta[,pstat1_index,third_index] - post_beta[,pstat1_index,first_index], 
  probs = c(0.025, 0.5, 0.975))
diff_log_count
exp(diff_log_count)
```

Plot posterior standard deviation.

```{r posterior_sigma}
p2 = plot(obj, type = "sigma") + 
  ggtitle("Marker Standard Deviation"~sigma) +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(ncol = 1)) +
  scale_color_manual(values=c("#5DA5DA", "#FAA43A", "#F17CB0"))
p2
```

Plot posterior correlations.

```{r posterior_cor}
plist = plot(obj, type = "Cor")
plist
```

Assess correlation uncertainity.

```{r correlation_uncertainty}
marker_pair = c("pSTAT3","pSTAT5")
Cor = rstan::extract(obj$fit_mcmc, pars = "Cor")[[1]]
Cor_term = rstan::extract(obj$fit_mcmc, pars = "Cor_term")[[1]]
Cor_diff = Cor_term - Cor
tb_cor = Cor_diff[,
                  which(protein_names == marker_pair[1]),
                  which(protein_names == marker_pair[2])] %>% as.tibble
tb_cor %<>% mutate(
  side = if_else(tb_cor$value > 0, 
                 true = paste0("positive (", 100*mean(tb_cor$value > 0), "%)"),
                 false = paste0("negative (", 100*mean(tb_cor$value <= 0), "%)"))
)
ggplot(tb_cor, aes(value, fill = side)) + 
  geom_histogram(bins = 50, alpha = 0.7) +
  xlab(paste0("Cor_term(", paste(marker_pair, collapse = ", "),")" )) +
  ggtitle("Posterior Distribution") + 
  scale_fill_manual(values = c("#E46726")) #"#6D9EC1"
```

Check if overall correlation structure changes between conditions.

```{r compare_covariance}
value = sapply(1:nrow(Cor_diff), function(i) {
  mask = which(upper.tri(Cor_diff[i,,]), arr.ind = T)
  cord = Cor_diff[i,,]
  mean(cord[lower.tri(cord)] > 0)
})
tb_cor = tibble(value = value)
tb_cor %<>% mutate(
  side = if_else(tb_cor$value > 0.5, 
                 true = paste0("> 1/2 (", 100*mean(tb_cor$value > 0.5), "%)"),
                 false = paste0("<= 1/2 (", 100*mean(tb_cor$value <= 0.5), "%)"))
)
p_global = ggplot(tb_cor, aes(value, fill = side)) + 
  geom_histogram(bins = 25, alpha = 0.7) +
  ggtitle(expression("Overall P(Corr"~Omega~"(3rd) > Corr"~Omega~"(1st))")) +
  scale_fill_manual(values = c("#E46726")) + #"#6D9EC1"
  theme(legend.position = "bottom") +
  xlab("probability")
p_global
```

Plot differential correlations.

```{r plot_differential_cor}
cor_increase = apply(X = Cor_diff, MARGIN = c(2,3), FUN = function(x) mean(x > 0))
colnames(cor_increase) = rownames(cor_increase) = protein_names
p_local = ggcorrplot(cor_increase, hc.order = TRUE, type = "lower",
           outline.col = "lightgray",
           colors = c("#6D9EC1", "white", "#E46726")) +
  ggtitle(expression("P(Corr"~Omega~"(3rd) > Corr"~Omega~"(1st))")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_gradient2(limit = c(0, 1), midpoint = 0.5,
                       low = "#6D9EC1", mid =  "white", high = "#E46726",
                       name = "probability")
p_local
```

Plot graph with edges at least 95% probability of larger correlation.

```{r plot_graph}
plot_correlation_graph = function(lambda) {
  graph = apply(Cor_diff, c(2,3), function(x) mean(x>0))
  diag(graph) = 0
  graph[upper.tri(graph)] = 0
  ind = which(graph > 0, arr.ind = T)
  tb_graph = tibble(
    from = protein_names[ind[,1]],
    to = protein_names[ind[,2]],
    prob = graph[ind]
  )
  tb_graph %<>% dplyr::filter(prob > lambda)
  tb_graph
  bayesFDR = sum(1-tb_graph$prob)/nrow(tb_graph)*100
  bayesFDR
  set.seed(0xdada)
  ig = graph_from_data_frame(tb_graph, directed = FALSE)
  ggplot(ggnetwork(ig, layout = "circle"), aes(x, y, xend = xend, yend = yend)) + 
    geom_edges(color = "black", size = 1) +
    geom_nodes(color = "black", size = 20) + 
    geom_nodetext(aes(label = vertex.names), 
                  color = "white", size = 3, fontface = "bold") +
    xlim(c(-0.1, 1.1)) +
    ylim(c(-0.1, 1.1)) + 
    ggtitle(paste0("Posterior Expected FDR: ", 
                   round(bayesFDR, digits = 1),"%")) +
    theme_blank() +
    theme(plot.title = element_text(hjust = 0.5))
}
plot_correlation_graph(lambda = 0.8)
```

Combine plot for paper.

```{r compbine_plot}
pall = plot_grid(
  p1, p2, 
  plist[[1]] + ggtitle(expression("Marker Corr"~Omega~"(1st trimester)")),
  plist[[2]] + ggtitle(expression("Marker Corr"~Omega~"(3rd trimester)")),
  p_global, p_local, 
  rel_heights = c(0.38,0.31,0.31),
  nrow = 3, labels = "AUTO"
)
ggsave(plot = pall, 
       filename = "posterior_summary_plmm.pdf", 
       width = 8, height = 11)
```

# Goodness of Fit

Define a test statistics and compare observed value with posterior predictive distribution. 

Predictive distribution marginalized over cell random effects.

```{r marginal_predicted_counts}
stan_pars = rstan::extract(obj$fit_mcmc, pars = c("beta",
                                                  "sigma","sigma_term","sigma_donor",
                                                  "Cor","Cor_term","Cor_donor"))
condition = "term"
term = df_samples_subset %>%
    pull(condition) %>%
    as.factor() %>%
    as.integer()
# kth posterior draw
sample_y_hat = function(k = 1) {
  lapply(1:2, function(cond) {
    n_cells_cond = table(term)[cond]
    beta = stan_pars$beta[k,,]
    mu = rep(0, length(protein_names))
    beta_rep = sapply(beta[,cond], rep, n_cells_cond)
    sigma = stan_pars$sigma[k,]
    Cor = stan_pars$Cor[k,,]
    Cov = diag(sigma) %*% Cor %*% diag(sigma)
    b = mvrnorm(n = n_cells_cond, mu, Cov)
    sigma_donor = stan_pars$sigma_donor[k,]
    Cor_donor = stan_pars$Cor_donor[k,,]
    Cov_donor = diag(sigma_donor) %*% Cor_donor %*% diag(sigma_donor)
    b_donor = mvrnorm(n = n_cells_cond, mu, Cov_donor)
    count = exp(beta_rep + b + b_donor)
    count = matrix(rpois(length(count), count),
                   nrow = nrow(count), 
                   ncol = ncol(count))
    count %<>% as.tibble
    names(count) = protein_names
    count %<>% add_column(term  = obj$conditions[cond])
    count
  }) %>% bind_rows()
}
Y_hat = sample_y_hat(k = 1)
Y_hat %>% 
  group_by(term) %>% 
  summarize_at(protein_names, median)
df_samples_subset %>% 
  group_by(term) %>% 
  summarize_at(protein_names, median)
```

Check if we can model pSTAT, pSTAT3, and pSTAT5 bright cells.

```{r goodness_of_fit_a}
gof = function(df, test_stat) {
  tfm = function(x) asinh(x/5)
  df_tfm = df %>% mutate_at(protein_names, tfm)
  df_median = df_tfm %>% summarize_at(protein_names, median)
  tibble(
    term = c(
      "1st trimester", 
      "3rd trimester"
      ),
    statistic = c(
      test_stat(df_tfm %>% filter(term == "1st trimester"), df_median), 
      test_stat(df_tfm %>% filter(term == "3rd trimester"), df_median)
      )
  )
}
test_stat_a = function(df_tfm, df_median) {
  mean(df_tfm$pSTAT1 > df_median$pSTAT1 & 
       df_tfm$pSTAT3 > df_median$pSTAT3 & 
       df_tfm$pSTAT5 > df_median$pSTAT5) * 100
}
gof_obsv_a = gof(df_samples_subset, test_stat_a)
gof_pred_a = mclapply(1:dim(stan_pars$beta)[1], function(k) gof(sample_y_hat(k), test_stat_a),
                      mc.cores = ncores) %>% bind_rows()
gof_obsv_a %<>% mutate(subset = "Cell Subset A")
gof_pred_a %<>% mutate(subset = "Cell Subset A")
```

Check if we can model pSTAT bright, and pSTAT3 and pSTAT5 dim cells.

```{r goodness_of_fit_b}
test_stat_b = function(df_tfm, df_median) {
  mean(df_tfm$pSTAT1 > df_median$pSTAT1 & 
       df_tfm$pSTAT3 < df_median$pSTAT3 & 
       df_tfm$pSTAT5 < df_median$pSTAT5) * 100
}
gof_obsv_b = gof(df_samples_subset, test_stat_b)
gof_pred_b = mclapply(1:dim(stan_pars$beta)[1], function(k) gof(sample_y_hat(k), test_stat_b),
                      mc.cores = ncores) %>% bind_rows()
gof_obsv_b %<>% mutate(subset = "Cell Subset B")
gof_pred_b %<>% mutate(subset = "Cell Subset B")
```

Check if we can model zero pERK1_2 and pMAPKAPK bright cells.

```{r goodness_of_fit_c}
test_stat_c = function(df_tfm, df_median) {
  mean(df_tfm$pERK1_2  == 0 & 
       df_tfm$pMAPKAPK > df_median$pMAPKAPK) * 100
}
gof_obsv_c = gof(df_samples_subset, test_stat_c)
gof_pred_c = mclapply(1:dim(stan_pars$beta)[1], function(k) gof(sample_y_hat(k), test_stat_c),
                      mc.cores = ncores) %>% bind_rows()
gof_obsv_c %<>% mutate(subset = "Cell Subset C")
gof_pred_c %<>% mutate(subset = "Cell Subset C")
```

Check if we can model nonzero pERK1_2 and pMAPKAPK bright cells.

```{r goodness_of_fit_d}
test_stat_d = function(df_tfm, df_median) {
  mean(df_tfm$pERK1_2  > 0 & 
       df_tfm$pMAPKAPK > df_median$pMAPKAPK
       ) * 100
}
gof_obsv_d = gof(df_samples_subset, test_stat_d)
gof_pred_d = mclapply(1:dim(stan_pars$beta)[1], function(k) gof(sample_y_hat(k), test_stat_d),
                      mc.cores = ncores) %>% bind_rows()
gof_obsv_d %<>% mutate(subset = "Cell Subset D")
gof_pred_d %<>% mutate(subset = "Cell Subset D")
```

Combined plot for paper.

```{r goodness_of_fit_combined}
# combine observed statistic
gof_obsv_all = bind_rows(gof_obsv_a, gof_obsv_b, gof_obsv_c, gof_obsv_d)
# combined predicted statistic
gof_pred_all = bind_rows(gof_pred_a, gof_pred_b, gof_pred_c, gof_pred_d)
# plot everything
ggplot(gof_pred_all, aes(statistic, fill = term)) + 
  geom_histogram(bins = 40, position = "identity", alpha = 0.5) +
  geom_vline(data = gof_obsv_all, linetype = "dashed", size = 1, 
             aes(xintercept = statistic, color = term)) +
  scale_fill_few() + 
  scale_color_few() + 
  xlab("test statistic (percentage)") + 
  facet_wrap(~subset, scales = "free") + 
  theme(legend.position="bottom")
ggsave(filename = "goodness_of_fit.pdf", 
       width = 8, height = 5)
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```

# References {.unnumbered}
