---
title: "Poisson Log-Normal Mixed Model"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

Reanalysis of mass cytometry data from @aghaeepour2017immune using the Poisson Log-Normal Mixed Model.

# Prerequisites

Parse input parameters.

```{r parse_input_parameters}
ncells = Inf
seed = 0xdada
ncores = 8
zenodo_url = "https://zenodo.org/record/2652578/files/"
cytof_data = "se_aghaeepour2017immune.Rdata"
prefit = paste0("cytoeffect_plmm_ncells_",ncells,".Rdata")
prefit
```

Install packages.

```{r install_packages, warning=FALSE, message=FALSE, eval=FALSE}
pkgs_needed = c("devtools","tidyverse","magrittr","SummarizedExperiment",
                "ggthemes","cowplot","RColorBrewer","broom","hexbin",
                "intergraph","igraph","ggnetwork","ggcorrplot","MASS",
                "parallel","dplyr")
letsinstall = setdiff(pkgs_needed, installed.packages())
if (length(letsinstall) > 0) {
  BiocManager::install(letsinstall)
}
devtools::install_github("ChristofSeiler/cytoeffect")
```

Load packages.

```{r load_packages}
library("cytoeffect")
library("tidyverse")
library("magrittr")
library("SummarizedExperiment")
library("ggthemes")
library("cowplot")
library("RColorBrewer")
library("broom")
library("intergraph")
library("igraph")
library("ggnetwork")
library("ggcorrplot")
library("MASS")
library("parallel")
library("dplyr")
theme_set(theme_few())
```

# Load Data

Download preprocessed data from Zenodo.

```{r download_zenodo}
rdata_filenames = c(cytof_data, prefit)
for(filename in rdata_filenames)
    download.file(url = paste0(zenodo_url, filename), 
                  destfile = filename,
                  mode = "wb")
```

Load ``SummarizedExperiment`` object from CytoGLMM workflow.

```{r load_summarized_experiment}
load(cytof_data)
exprs = assay(se_aghaeepour2017immune)
sample_info = rowData(se_aghaeepour2017immune)
sample_info_names = names(sample_info)
df_samples = cbind(as.data.frame(exprs), as.data.frame(sample_info))
df_samples %<>% as.tibble
protein_names = colData(se_aghaeepour2017immune) %>% 
  as.data.frame %>% 
  dplyr::filter(type == "function") %>%
  .$protein_name
gate_protein_names = colData(se_aghaeepour2017immune) %>% 
  as.data.frame %>% 
  dplyr::filter(type == "phenotype") %>%
  .$protein_name
```

# Fit Model

Tally cell count.

```{r cell_count}
df_samples %>% group_by(term,celltype,donor) %>% tally %>% arrange(n)
df_samples %>% group_by(term,celltype,donor) %>% tally %>% arrange(desc(n))
```

Subset to NK cells.

```{r one_donor_one_cell_subset}
df_samples_subset = df_samples %>% dplyr::filter(celltype == "NK")
df_samples_subset %<>% dplyr::select(protein_names,gate_protein_names,sample_info_names)
```

Subsample cells to a maximum number of cells per donor.

```{r subsample_cells}
if(nrow(df_samples_subset) > ncells) {
  print(paste("subsampled to",ncells,"per donor"))
  set.seed(seed)
  # subsample depending on max cell count
  df_count = df_samples_subset %>% group_by(donor) %>% tally() %>%
    mutate(nnew = ifelse(n > ncells,ncells,n))
  # create table with a data frame in one column
  df_nested = df_samples_subset %>% group_by(donor) %>% nest() %>%
    left_join(df_count,by = "donor")
  # subsample per donor
  df_samples_subset = df_nested %>%
    mutate(samp = map2(data, nnew, sample_n)) %>%
    dplyr::select(donor, samp) %>%
    unnest()
} else {
  print("no subsampling done")
}
```

Tally cell count.

```{r cell_count_subsample}
df_samples_subset %>% group_by(term,celltype,donor) %>% tally %>% 
  arrange(n)
df_samples_subset %>% group_by(term,celltype,donor) %>% tally %>% 
  arrange(desc(n))
```

## HMC Sampling

Sample from posterior distribution (initialization details in paper).

```{r set_reference}
df_samples_subset$term %<>% factor(levels = c("1st trimester",
                                              "3rd trimester"))
```

```{r poisson_sampling}
if(file.exists(prefit)) {
  load(file = prefit)
} else {
  obj = cytoeffect::poisson_lognormal(df_samples_subset, protein_names, 
                                      condition = "term", group = "donor",
                                      r_cell = 3, r_donor = 3,
                                      iter = 325, warmup = 200,
                                      num_chains = ncores)
  save(obj,file = prefit)
}
```

## HMC Diagnostics

Postprocessing of posterior samples. Traceplot of posterior samples.

```{r post_sampling, fig.wide=TRUE}
pars_str = "beta"
rstan::traceplot(obj$fit_mcmc, inc_warmup = FALSE, pars = pars_str)
# test case
stan_pars = rstan::extract(obj$fit_mcmc,
                           pars = c("sigma_donor", "Q_donor", "Cor_donor"))
k = 100
sigma = stan_pars$sigma_donor[k,]
Q = stan_pars$Q_donor[k,,]
cor = cov2cor(Q %*% diag(sigma^2) %*% t(Q))
range(cor - stan_pars$Cor_donor[k,,])
```

Some more MCMC diagnostics. According to empirically findings, Rhat > 1.1 is usually indicative of problems in the fit.

```{r mcmc_diagnostics}
pars = c("beta",
         "sigma","sigma_term","sigma_donor",
         "Cor","Cor_term","Cor_donor",
         "b_donor")
tb = summary(obj$fit_mcmc, 
             pars = pars)$summary %>% 
  as.tibble(rownames = "pars", .before = 1) %>% 
  dplyr::select(pars, n_eff, Rhat)
tb %<>% na.omit() # Stan fills upper triangle with zeros
tb %>% arrange(n_eff)
tb %>% arrange(desc(Rhat))
tb %>% summarize(min = min(n_eff), max = max(n_eff))
tb %>% summarize(min = min(Rhat), max = max(Rhat))
```

# Results

Plot posterior regression coefficients.

```{r plot_beta}
p1 = plot(obj, type = "beta") + 
  ggtitle(expression("Fixed Effects"~beta)) +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(ncol = 1)) + 
  scale_color_few()
p1
plot(obj, type = "beta") + 
  facet_wrap(~condition, scales = "free_x") +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(ncol = 1)) + 
  scale_color_few()
```

Extract expected count difference for pSTAT1.

```{r fixed_effects_pSTAT1}
post_beta = rstan::extract(obj$fit_mcmc, pars = "beta")[[1]]
first_index = which(levels(pull(obj$df_samples_subset, obj$condition)) 
                    == "1st trimester")
third_index = which(levels(pull(obj$df_samples_subset, obj$condition)) 
                    == "3rd trimester")
pstat1_index = which(obj$protein_names == "pSTAT1")
first_log_count = quantile(post_beta[,pstat1_index,first_index], 
                           probs = c(0.025, 0.5, 0.975))
first_log_count
exp(first_log_count)
third_log_count = quantile(post_beta[,pstat1_index,third_index], 
                           probs = c(0.025, 0.5, 0.975))
third_log_count
exp(third_log_count)
diff_log_count = quantile(
  post_beta[,pstat1_index,third_index] - post_beta[,pstat1_index,first_index], 
  probs = c(0.025, 0.5, 0.975))
diff_log_count
exp(diff_log_count)
```

Posterior multivariate pairs plot.

```{r posterior_pair_plot}
pSTAT1_index = which(obj$protein_names == "pSTAT1")
pSTAT3_index = which(obj$protein_names == "pSTAT3")
pSTAT5_index = which(obj$protein_names == "pSTAT5")
post_beta = rstan::extract(obj$fit_mcmc, pars = "beta")[[1]]
tb_log_count = bind_rows(
  tibble(
    term = levels(pull(obj$df_samples_subset, obj$condition))[1],
    pSTAT1 = post_beta[,pSTAT1_index,1],
    pSTAT3 = post_beta[,pSTAT3_index,1],
    pSTAT5 = post_beta[,pSTAT5_index,1]
  ),
  tibble(
    term = levels(pull(obj$df_samples_subset, obj$condition))[2],
    pSTAT1 = post_beta[,pSTAT1_index,2],
    pSTAT3 = post_beta[,pSTAT3_index,2],
    pSTAT5 = post_beta[,pSTAT5_index,2]
  )
)
plot_diag = function(marker) {
  ggplot(tb_log_count, aes_string(marker, fill = "term")) + 
    geom_histogram(bins = 40, position = "identity", alpha = 0.5) +
    scale_fill_few()
}
plot_off_diag = function(marker1, marker2) {
  ggplot(tb_log_count, aes_string(marker1, marker2, color = "term")) +
    geom_density2d() + 
    scale_color_few()
}
ppair = plot_grid(
  plot_diag("pSTAT1") + theme(legend.position = "none"),
  NULL, 
  NULL,
  plot_off_diag("pSTAT1","pSTAT3") + theme(legend.position = "none"), 
  plot_diag("pSTAT3") + theme(legend.position = "none"), 
  NULL,
  plot_off_diag("pSTAT1","pSTAT5") + theme(legend.position = "none"), 
  plot_off_diag("pSTAT3","pSTAT5") + theme(legend.position = "none"), 
  plot_diag("pSTAT5") + theme(legend.position = "none"),
  ncol = 3
)
plot_grid(ppair,
          get_legend(plot_diag("pSTAT1") + theme(legend.position = "bottom")),
          ncol = 1,
          rel_heights = c(1, .1))
ggsave(filename = "posterior_multivariate_plmm.pdf", width = 8, height = 6)
```

Plot posterior standard deviation.

```{r posterior_sigma}
p2 = plot(obj, type = "sigma") + 
  ggtitle("Marker Standard Deviation"~sigma) +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(ncol = 1)) +
  scale_color_manual(values=c("#5DA5DA", "#FAA43A", "#F17CB0"))
p2
```

Plot posterior correlations.

```{r posterior_cor}
plist = plot(obj, type = "Cor")
plist
```

Multivariate posterior MDS plots.

```{r multivariate_posterior}
cytoeffect::plot_mds(obj)
ggsave(filename = "posterior_mds_latent_variable_mu_scaled.pdf", width = 10, height = 5)
cytoeffect::plot_mds(obj, asp = FALSE)
ggsave(filename = "posterior_mds_latent_variable_mu_unscaled.pdf", width = 10, height = 5)
cytoeffect::plot_mds(obj, asp = FALSE, show_donors = FALSE)
ggsave(filename = "posterior_mds_latent_variable_mu_unscaled_no_donors.pdf", 
       width = 10, height = 5)
```

Pairwise correlation change between conditions.

```{r correlation_uncertainty}
marker_pair = c("pSTAT3","pSTAT5")
Cor = rstan::extract(obj$fit_mcmc, pars = "Cor")[[1]]
Cor_term = rstan::extract(obj$fit_mcmc, pars = "Cor_term")[[1]]
Cor_diff = Cor_term - Cor
tb_cor = Cor_diff[,
                  which(obj$protein_names == marker_pair[1]),
                  which(obj$protein_names == marker_pair[2])] %>% as.tibble
tb_cor %<>% mutate(
  side = if_else(tb_cor$value > 0, 
                 true = paste0("positive (", 100*mean(tb_cor$value > 0), "%)"),
                 false = paste0("negative (", 100*mean(tb_cor$value <= 0), "%)"))
)
# keep colors consistent
if(mean(tb_cor$value > 0) == 1) {
  fill_colors = "#E46726"
} else {
  fill_colors = c("#6D9EC1","#E46726")
}
ggplot(tb_cor, aes(value, fill = side)) + 
  geom_histogram(bins = 50, alpha = 0.7) +
  xlab(paste0("Cor_term(", paste(marker_pair, collapse = ", "),")" )) +
  ggtitle("Posterior Distribution") + 
  scale_fill_manual(values = fill_colors)
```

Check if overall correlation structure changes between conditions.

```{r compare_covariance}
value = sapply(1:nrow(Cor_diff), function(i) {
  mask = which(upper.tri(Cor_diff[i,,]), arr.ind = T)
  cord = Cor_diff[i,,]
  mean(cord[lower.tri(cord)] > 0)
})
tb_cor = tibble(value = value)
tb_cor %<>% mutate(
  side = if_else(tb_cor$value > 0.5, 
                 true = paste0("> 1/2 (", 100*mean(tb_cor$value > 0.5), "%)"),
                 false = paste0("<= 1/2 (", 100*mean(tb_cor$value <= 0.5), "%)"))
)
p_global = ggplot(tb_cor, aes(value, fill = side)) + 
  geom_histogram(bins = 25, alpha = 0.7) +
  ggtitle(expression("Overall P(Corr"~Omega~"(3rd) > Corr"~Omega~"(1st))")) +
  scale_fill_manual(values = fill_colors) +
  theme(legend.position = "bottom") +
  xlab("probability")
p_global
```

Plot differential correlations.

```{r plot_differential_cor}
cor_increase = apply(X = Cor_diff, MARGIN = c(2,3), FUN = function(x) mean(x > 0))
colnames(cor_increase) = rownames(cor_increase) = obj$protein_names
p_local = ggcorrplot(cor_increase, hc.order = TRUE, type = "lower",
           outline.col = "lightgray",
           colors = c("#6D9EC1", "white", "#E46726")) +
  ggtitle(expression("P(Corr"~Omega~"(3rd) > Corr"~Omega~"(1st))")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_gradient2(limit = c(0, 1), midpoint = 0.5,
                       low = "#6D9EC1", mid =  "white", high = "#E46726",
                       name = "probability")
p_local
```

Plot graph with edges at least 95% probability of larger correlation.

```{r plot_graph}
plot_correlation_graph = function(lambda) {
  graph = apply(Cor_diff, c(2,3), function(x) mean(x>0))
  diag(graph) = 0
  graph[upper.tri(graph)] = 0
  ind = which(graph > 0, arr.ind = T)
  tb_graph = tibble(
    from = obj$protein_names[ind[,1]],
    to = obj$protein_names[ind[,2]],
    prob = graph[ind]
  )
  tb_graph %<>% dplyr::filter(prob > lambda)
  tb_graph
  bayesFDR = sum(1-tb_graph$prob)/nrow(tb_graph)*100
  bayesFDR
  set.seed(0xdada)
  ig = graph_from_data_frame(tb_graph, directed = FALSE)
  ggplot(ggnetwork(ig, layout = "circle"), aes(x, y, xend = xend, yend = yend)) + 
    geom_edges(color = "black", size = 1) +
    geom_nodes(color = "black", size = 20) + 
    geom_nodetext(aes(label = vertex.names), 
                  color = "white", size = 3, fontface = "bold") +
    xlim(c(-0.1, 1.1)) +
    ylim(c(-0.1, 1.1)) + 
    ggtitle(paste0("Posterior Expected FDR: ", 
                   round(bayesFDR, digits = 1),"%")) +
    theme_blank() +
    theme(plot.title = element_text(hjust = 0.5))
}
plot_correlation_graph(lambda = 0.8)
```

Combine plot for paper.

```{r compbine_plot}
pall = plot_grid(
  p1, p2, 
  plist[[1]] + ggtitle(expression("Marker Corr"~Omega~"(1st trimester)")),
  plist[[2]] + ggtitle(expression("Marker Corr"~Omega~"(3rd trimester)")),
  p_global, p_local, 
  rel_heights = c(0.38,0.31,0.31),
  nrow = 3, labels = "AUTO"
)
ggsave(plot = pall, 
       filename = "posterior_summary_plmm.pdf", 
       width = 8, height = 11)
```

# Goodness of Fit

Define a test statistics and compare observed value with posterior predictive distribution.

```{r marginal_predicted_counts}
sample_y_hat = function(k) {
  tb = posterior_predictive_log_lambda(obj, k = 1, show_donors = TRUE)
  lambda_matrix = tb %>% 
    dplyr::select(obj$protein_names) %>% 
    as.matrix %>% 
    exp
  Y_hat = matrix(rpois(length(lambda_matrix), lambda_matrix),
                   nrow = nrow(lambda_matrix), 
                   ncol = ncol(lambda_matrix))
  tb[1:length(obj$protein_names)] = Y_hat
  tb
}
Y_hat = sample_y_hat(k = 1)
Y_hat %>% 
  group_by(term) %>% 
  summarize_at(obj$protein_names, median)
obj$df_samples_subset %>% 
  group_by(term) %>% 
  summarize_at(obj$protein_names, median)
```

Four checks:

* Subset A: Check if we can model pSTAT1, pSTAT3, and pSTAT5 bright cells
* Subset B: Check if we can model pSTAT1 bright, and pSTAT3 and pSTAT5 dim cells
* Subset C: Check if we can model zero pERK1_2 and pMAPKAPK bright cells
* Subset D: Check if we can model nonzero pERK1_2 and pMAPKAPK bright cells

```{r goodness_of_fit}
set.seed(0xdada)
pSTAT1_median = median(obj$df_samples_subset$pSTAT1)
pSTAT3_median = median(obj$df_samples_subset$pSTAT3)
pSTAT5_median = median(obj$df_samples_subset$pSTAT5)
pMAPKAPK_median = median(obj$df_samples_subset$pMAPKAPK)
gof = function(tb) {
  tb %>% 
    group_by(term, donor) %>% 
    summarize(
      pSTAT_bright = mean(pSTAT1 > pSTAT1_median & 
                          pSTAT3 > pSTAT3_median & 
                          pSTAT5 > pSTAT5_median),
      pSTAT1_bright_35_dim = mean(pSTAT1 > pSTAT1_median & 
                                  pSTAT3 < pSTAT3_median & 
                                  pSTAT5 < pSTAT5_median),
      pERK1_2_zero = mean(pERK1_2  == 0),
      pMAPKAPK_nonzero = mean(pMAPKAPK > pMAPKAPK_median)
      )
}
n_iter = nrow(rstan::extract(obj$fit_mcmc, pars = "beta")[[1]])
gof_pred = mclapply(1:n_iter, function(k) gof(sample_y_hat(k)), 
                    mc.cores = ncores, mc.set.seed = FALSE) %>% bind_rows()
gof_obsv = gof(obj$df_samples_subset)
```

Observed test statistic (vertical dashed line) should be in the bulk of the distribution.

```{r plot_goodness_of_fit}
gof_pred_long = gof_pred %>% 
  gather(subset, statistic, -term, -donor)
gof_obsv_long = gof_obsv %>% 
  gather(subset, statistic, -term, -donor)
ggplot(gof_pred_long, aes(statistic, fill = term)) + 
  geom_histogram(bins = 40, position = "identity", alpha = 0.5) +
  geom_point(data = gof_obsv_long, 
             position = position_jitter(width = 0, height = 50), alpha = 0.8,
             aes(y = 0, x = statistic, color = term)) +
  scale_fill_few() + 
  scale_color_few() + 
  xlab("test statistic") + 
  facet_wrap(~subset, scales = "free") + 
  theme(legend.position="bottom") +
  ggtitle("All Donors")
ggsave(filename = "goodness_of_fit.pdf", 
       width = 8, height = 5)
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```

# References {.unnumbered}
